---
title: "`r params$doctitle`"
params:
  doctitle: "Process Modeling and Deployment"
author: 
- João Caldeira, [jcppc@iscte-iul.pt](mailto:jcppc@iscte-iul.pt), ISCTE-IUL^[Invited Assistant Professor. https://ciencia.iscte-iul.pt/authors/joao-caldeira/cv]
#^[LinkedIn. https://www.linkedin.com/in/joao-carlos-caldeira/]
- Fernando Brito e Abreu, [fba@iscte-iul.pt](mailto:fba@iscte-iul.pt), ISCTE-IUL^[Associate Professor. https://ciencia.iscte-iul.pt/authors/fernando-brito-e-abreu/] 
# name: João Caldeira^[Invited Professor. https://www.linkedin.com/in/joao-carlos-caldeira/]
# email: jcppc@iscte-iul.pt
# affiliation: ISCTE-IUL
date: "`r format(Sys.time(), '%d %B %Y')`"
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{\includegraphics[width=3cm]{iscte.png}}
- \fancyfoot[L]{Copyright @ `r format(Sys.time(), '%Y')` Department of Information Science and Technology}
- \fancyfoot[C]{}
- \fancyfoot[R]{\thepage}
#abstract: This is the abstract.
output:
  html_document: 
    toc: yes
    toc_float: yes
    highlight: espresso
    theme: spacelab
  html_notebook: 
    toc: yes
    toc_float: yes
    number_sections: yes
    fig_caption: yes
    highlight: espresso
    theme: spacelab
  pdf_document: 
    toc: yes
    fig_caption: yes
    number_sections: yes
    highlight: espresso
  word_document:
    toc: yes
    toc_depth: '3'
subtitle: Process Data Analysis - Insights/Inputs for Optimized Process Simulation
tags:
- process modeling
- process simulation
- distribution assessment
---

\newpage

# Environment Setup

## Install R

Install **R** from here: [https://cran.r-project.org/](https://cran.r-project.org/).

## Install R Studio Desktop

Install **R Studio Desktop** from here: [https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download).


## Install Required Libraries
```{r, eval=FALSE, echo=TRUE}
install.packages("readxl")
install.packages("fitdistrplus")
install.packages("stats")
install.packages("actuar")
install.packages("vtable")
install.packages("data.table")
install.packages("dplyr")
install.packages("vioplot")
install.packages("bupaR")
install.packages("processanimateR")
```

## Load Required Libraries
```{r, warning=FALSE,message=FALSE,echo=TRUE}
library(readxl)
library(fitdistrplus)
library(stats)
library(actuar)
library(vtable)
library(data.table)
library(dplyr)
library(vioplot)
library(bupaR)
library(processanimateR)
```

```{r echo=FALSE}
options(scipen = 999)
palette <- c("#14bfdd", "#80b3d2", "#040404", "#d7d9ce")
```

\newpage

# Dataset Preparation
## Loading the Dataset
This **filename** should match your data location. The log file should have at least the columns named **Case ID**, **Activity**, **Start Time**, **End Time**.
```{r}
dataset <- read_excel("C:/Users/Usuario/Desktop/Pastas/Escola/Universidade/Mestrado/1º Semestre/Modelação e Implementação de Processos/Trabalho Grupo/2ª Parte/G17.xlsx")
#colnames(dataset)[which(names(dataset) == "case")] <- "Case ID"
#colnames(dataset)[which(names(dataset) == "event")] <- "Activity"
#colnames(dataset)[which(names(dataset) == "resource")] <- "Resource"
colnames(dataset)[which(names(dataset) == "Start Timestamp")] <- "Start Time"
colnames(dataset)[which(names(dataset) == "Complete Timestamp")] <- "End Time"
```

## Formatting Date/Time columns
The **Start Time** and **End Time** columns should be formatted accordingly with the proper *Date/Time* mask present in the log file.
```{r}
dataset[['Start Time']] <- as.POSIXct(dataset$`Start Time`, format = "%Y-%m-%d %H:%M:%S")
dataset[['End Time']] <- as.POSIXct(dataset$`End Time`, format = "%Y-%m-%d %H:%M:%S")
```


## Viewing the Dataset
```{r}
dataset
```

\newpage
## Filtering Only First Event of Each Case/Instance
```{r tidy=TRUE, tidy.opts=list(width.cutoff=65)}
#This is needed to compute the arrival time between instances/cases
filtered <- dataset[!duplicated(dataset$`Case ID`),] 
filtered
```

## Ordering by Start Date/Time
```{r}
ordered <- filtered[order(as.POSIXct(filtered$`Start Time`), decreasing = FALSE),]
ordered
```

\newpage
## Computing Arrival Interval Between Cases/Instances
```{r tidy=TRUE, tidy.opts=list(width.cutoff=65)}
ordered <- data.table(ordered)
instances <- ordered[, Interval := difftime(as.POSIXct(ordered$`Start Time`), shift(as.POSIXct(ordered$`Start Time`), fill=ordered$`Start Time`[1L]), units="mins")]
instances <- instances[-c(1),]
instances
summary(as.numeric(instances$Interval))
sd(as.numeric(instances$Interval)) #Standard Deviation
var(as.numeric(instances$Interval)) #Variance
```


## Computing Activities Elapsed Time
```{r tidy=TRUE, tidy.opts=list(width.cutoff=65)}
dataset[['Elapsed Time']] <- as.difftime(difftime(dataset[['End Time']], dataset[['Start Time']], units="mins"))
model <- dataset
datasetActivities <- dataset
datasetActivities[,-c(1)][datasetActivities[,-c(1)]==0] <- NA
datasetActivities <- datasetActivities[complete.cases(datasetActivities),]
datasetActivities #[,c(1:6,7)] #Note: Viewing only some columns with changed order
summary(as.numeric(datasetActivities$`Elapsed Time`))
sd(as.numeric(datasetActivities$`Elapsed Time`)) #Standard Deviation
var(as.numeric(datasetActivities$`Elapsed Time`)) #Variance

```


## Computing Maximum and Minimum Hours of Work for Roles
```{r}
datasetTemp <- dataset
datasetTemp[['Start Hours']] <- as.POSIXct(strftime(datasetTemp$`Start Time`, format="%H:%M:%S"), format = "%H:%M:%S")
datasetTemp[['End Hours']] <- as.POSIXct(strftime(datasetTemp$`End Time`, format="%H:%M:%S"), format = "%H:%M:%S")

labourHours <- datasetTemp %>%
  group_by(Role) %>%
  summarise(
    "Minimum Hour" = min(`Start Hours`, na.rm = T),
    "Maximum Hour" = max(`End Hours`, na.rm = T)
  ) %>%
  arrange(Role)

labourHours[['Min. Hour']] <- format(labourHours$`Minimum Hour`, "%H:%M:%S")
labourHours[['Max. Hour']] <- format(labourHours$`Maximum Hour`, "%H:%M:%S")
labourHours[,-c(2,3)]

#par(mfrow=c(1,2))
proportion <- table(dataset$Role)/nrow(dataset)

boxplot(datasetTemp$`Start Time` ~ datasetTemp$Role, width=proportion, ylab="", xlab="Roles", las=1, cex.axis = 0.5) 
vioplot(as.numeric(datasetTemp$`Start Time`) ~ datasetTemp$Role, ylab="", xlab="Roles", las=1, cex.axis = 0.5, col = palette[2], horizontal = FALSE)

boxplot(datasetTemp$`End Time` ~ datasetTemp$Role, width=proportion, ylab="", xlab="Roles", las=1, cex.axis = 0.5)
vioplot(as.numeric(datasetTemp$`End Time`) ~ datasetTemp$Role, ylab="", xlab="Roles", las=1, cex.axis = 0.5, col = palette[2], horizontal = FALSE)

```





\newpage
# Dataset Evaluation
## Instances Arrival Interval 
### Statistics
```{r}
instances.interval <- as.numeric(instances$Interval)
summary(instances.interval)
```

```{r tidy=TRUE, tidy.opts=list(width.cutoff=65)}
par(mfrow=c(2,2)) #Set a 2x2 matrix to plot 4 charts
hist(instances.interval, breaks = 20, main = "Histogram - Interval", xlab = "Interval")
hist(instances.interval, main = "Histogram - Interval (Zoom In)", breaks = 300, xlim = c(0,100), xlab = "Interval")
boxplot(instances.interval)
boxplot(instances.interval, ylim = c(0, 20)) #This is a scale reduced boxplot
```

\newpage
### Assess Distribution
```{r}
descdist(instances.interval, discrete = FALSE)
```

\newpage
### Fit/Compare the Distribution
```{r}
# Distributions from "stats" package: norm,lnorm,exp,pois,cauchy,gamma,logis,nbinom,
# geom, beta,weibull
# Distributions from "actuar" package: invgamma,llogis,invweibull,pareto1,pareto
distribution <- fitdist(instances.interval, "norm")
plot(distribution)
```

\newpage
## Activities Elapsed Time 
### Statistics
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
activities.duration <- st(dataset, group = "Activity", vars=c("Elapsed Time"), group.long= TRUE, out = "return")
activities.duration
```

\newpage
### Assess Distribution
```{r}
activities <- unique(dataset$Activity) # Get distinct activities names
```


```{r echo=FALSE}
for(i in 1:length(activities)) {
    cat(i,":",activities[i],"| ")
    if ( i %% 3 == 0 ) cat("\n")
}
```

```{r}
uniqueActivities = data.frame(ID=character(), Activity=character(), Value=integer())
#colnames(uniformActivities) <- c("ID", "Name", "Value")
for(i in 1:length(activities)) {
  activity <- filter(dataset, Activity == activities[i])
  if ( length(unique(activity$`Elapsed Time`)) == 1) { uniqueActivities <- rbind(uniqueActivities, data.frame(ID=i, Activity=activities[i],Value=unique(activity$`Elapsed Time`)));  cat("\n", "Activity - ",i," : " ,activities[i],"with uniform value of : ", unique(activity$`Elapsed Time`),"\n"); next; }
  descdist(as.numeric(activity$`Elapsed Time`), discrete = FALSE, graph = TRUE)
  cat("\n", "Activity - ",i," : " , activities[i], "\n")
}
```
### Activities with Unique Values
```{r}
uniqueActivities
```


\newpage
### Fit/Compare the Distribution

```{r echo=FALSE}
cat("\n", "Activity - " , activities[1], "\n")
```

```{r}
# Distributions from "stats" package: norm,lnorm,exp,pois,cauchy,gamma,logis,nbinom,
# geom,beta,weibull
# Distributions from "actuar" package: invgamma,llogis,invweibull,pareto1,pareto
# The potential distribution was obtained in previous step. eg: norm, exp, etc
# Repeat below 3 lines per each activity name/index trying a corresponding distribution
activity <- filter(dataset, Activity == activities[1])  # Activity index 1
distribution <- fitdist(as.numeric(activity$`Elapsed Time`), "exp")
plot(distribution)
```
\newpage

```{r echo=FALSE}
cat("\n", "Activity - " , activities[2], "\n")
```

```{r}
# Distributions from "stats" package: norm,lnorm,exp,pois,cauchy,gamma,logis,nbinom,
# geom,beta,weibull
# Distributions from "actuar" package: invgamma,llogis,invweibull,pareto1,pareto
# The potential distribution was obtained in previous step. eg: norm, exp, etc
# Repeat below 3 lines per each activity name/index trying a corresponding distribution
activity <- filter(dataset, Activity == activities[4]) # Activity index 4
distribution <- fitdist(as.numeric(activity$`Elapsed Time`), "norm")
plot(distribution)
```

\newpage
# Appendix A

## Probability Distributions (Examples)

```{r echo=FALSE}
observations <- 5000
bins.break <- observations*0.01
line.width <- 2
plot.matrix <- c(2,2) #Set a 2x2 matrix to plot 4 charts
```

### Normal Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rnorm(observations, mean = 0, sd = 1)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```

\newpage
### Exponential Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rexp(observations, rate = 1)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```

\newpage
### Chi-Square Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rchisq(observations, df = 5)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```


\newpage
### Logistic Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rlogis(observations)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```


\newpage
### Poisson Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rpois(observations, lambda = 15)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```


\newpage
### t-Student Distribution
```{r tidy=TRUE, tidy.opts=list(width.cutoff=70)}
set.seed(1) 
par(mfrow=plot.matrix) 
sample.dataset <- rt(observations, df = 10)
summary(sample.dataset)
hist(sample.dataset, main = "Histogram (Frequency)", breaks = bins.break, col = palette[1])
hist(sample.dataset, main = "Histogram (Density)", breaks = bins.break, col = palette[2], freq = FALSE)
lines(density(sample.dataset), lw = line.width, col = palette[3])
vioplot(sample.dataset, col = palette[2], horizontal = TRUE)
plot(density(sample.dataset), type="n", main ="")
polygon(density(sample.dataset), col=palette[4], border=palette[3])
rug(sample.dataset, col=palette[1])
```

## Viewing the Process Model
```{r}
include.in.render <- TRUE # Change this to FALSE to export to pdf (Knit to Pdf)
events <- model %>% #a data.frame with the information in the table above
    mutate(Status = "complete",
           Activity_Instance = 1:nrow(.)) %>%
    eventlog(
        case_id = "Case ID",
        activity_id = "Activity",
        activity_instance_id = "Activity_Instance",
        lifecycle_id = "Status",
        timestamp = "Start Time",
        resource_id = "Resource"
    )
```


```{r, eval=include.in.render, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
events %>% process_map(type = frequency("relative"), layout = layout_pm(fixed_positions = TRUE))
```


```{r, eval=include.in.render, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
events %>% process_map(performance(mean, "hours"))
```


```{r, eval=include.in.render, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
events %>%  precedence_matrix(type = "absolute")
```


```{r, eval=include.in.render, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70), fig.width=7}
events %>%  precedence_matrix(type = "absolute") %>% plot
```


```{r, eval=include.in.render, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=70)}
#events %>% resource_map()
animate_process(events, mode = "relative", jitter = 10, legend = "color",
                mapping = token_aes(size = token_scale(3), color = token_scale("Credit Request Cases/Instances", scale = "ordinal", 
                range = RColorBrewer::brewer.pal(12, "Paired"))))

```











